Modeling HEPES buffer with a charge-balance approach

### 2024-02-04 manuscript.Rmd
Import edits up to figure 2

```{r, echo=FALSE}
# NOTE: bookdown features (e.g. cross-references) used below
# This requires the use of a bookdown output format, e.g.
# bookdown::pdf_document2 or bookdown::html_document2
```

@@

It has been shown [@Ring2022] that $\mathrm{pH}$ in arbitrary fluids can be
understood based on the three fundamental principles of
electroneutrality, conservation of mass, and rules of dissociation as
devised from physical chemistry. It has also been shown [@Ring2022] that the
charge-balance approach allows definition of the $\mathrm{pH}$-dependent
buffer-capacity, and provides a definition of acid content as titratable
acidity, based on previous work by Kildeberg [@Kildeberg1983]. Recently
it has been reiterated [@Ring2023-charge] that it is absolutely necessary for this
modeling approach to take into account the strong ions, which are those
assumed to be completely dissociated at any $\mathrm{pH}$. In contrast to the
physico-chemical model of Stewart [@Stewart1978], the charge-balance
model (CBM, henceforth) has no separation between independent and
dependent variables. Under the CBM all that can be known about the
system is gathered in the expression of charge in an explicitly stated
chemical system. Consequently, if, as an example, $\mathrm{pH}$ is known and charge
is a known deterministic function of the chemical specification, inverse
inferences on unknown variables can, at least sometimes, be obtained by
root finding, since we know that the charge must be zero. In simple
systems, such as a single monovalent weak acid and strong ions, a single
unknown variable is explicitly estimable [@Ring2022]. The question
focused here is if CBM is rich enough to allow inverse inference in more
complex systems such as with multiprotic buffers. This question arose
under experiments with HEPES buffered sulfate containing fluids
[@Ring2023-acid] in which HEPES was modeled as a monovalent buffer with
pKa 7.5 [@Ring2010] since it was found that estimated $\mathrm{pH}$ was much too
low when observed $\mathrm{pH}$ was less than about 6. HEPES is a zwitterion with
formula 2-[4-(2-Hydroxyethyl)piperazin-1-yl]ethane-1-sulfonic acid and
structure as shown in \@ref(fig:hepes).

@@

Wikipedia informs [@WP:HEPES] that HEPES has two pK values, one at 3 and
one at 7.5, but the reference given [@Johnson2014] in Wikipedia states
that "3" comes from the sulfonate group and 7.5 from one of the
piperazine nitrogens.  At $20^\circ\mathrm{C}$ Good et al [@Good1966]
has pKa 3 and 7.55 without detailing which groups are at work, while
another study [@AlGhobashy2014] mentions that 3 and 7.5 both belongs to
piperazine amines and informs that the sulfonate group is dissociated
over "almost the whole $\mathrm{pH}$ range".  Furthermore, is has been stated that
the nitrogen at the $\mathrm{OH}$ end of HEPES is protonated in methanol
and the nitrogen in the sulfonate end is protonated in water
[@Sledz2010].

Hence it was not clear how best to model the $\mathrm{pH}$-dependent
charge on HEPES to estimate $\mathrm{pH}$ based on known chemistry.
Furthermore, it was desired to make clear if the $\mathrm{p}K$ values
for HEPES could be elucidated based on measured $\mathrm{pH}$. Herein,
these questions will be approached by first developing a model for the
mean charge on HEPES as a function of $\mathrm{pH}$ and applying this to
proposed values of set of underlying pKs using data obtained previously
[@Ring2023-acid]. Next, it will be examined if data generated with
specified $\mathrm{p}K$ values over extended or restricted range of
$\mathrm{pH}$ values allows recovery of these same $\mathrm{p}K$ values.
Accepting that $\mathrm{pH} = -1$ would hardly be realized in biological
fluid and, anyway would require dramatic ionic strengths invalidating
conventional understanding of $\mathrm{pH}$ modeling, the consequence
for modeling of assuming the first $\mathrm{p}K_1$ fixed at different
low values will be examined. In so far as the developed model for
$\mathrm{pH}$ dependent charge on HEPES implies a strict ordering of
$\mathrm{p}K$ values, as will be seen in the Methods section, the
consequence for recovery of known $\mathrm{p}K$ values of assuming
strict ordering will be assessed. The final model, thus developed, was
used with the original observed data to obtain best values for the
underlying $\mathrm{p}K$ values and the fit assessed. Finally, the
sensitivity of the final model to imposed noise in $\mathrm{SID}$,
$\mathrm{[HEPES]}$, or $\mathrm{pH}$ will be presented.

@@

```{r echo=FALSE}
# A dash! A dash! My kingdom for a dash!
# (The things that rmarkdown users have to do without
# \usepackage{mhchem})
```
$\mathrm{I: H^{+}N_1\text-HEP\text-N_4H^{+}SO_3H}$ $\sim \mathrm{p}K_{a1}$ on
sulfonic group

@@

So, we can see that $\mathrm{IV}$ is identified from total HEPES
concentration at a given $\mathrm{pH}$ if the three $\mathrm{p}K$ values
are known, and the derivation clearly requires that $\mathrm{p}K_{a1} <
\mathrm{p}K_{a2} < \mathrm{p}K_{a3}$. Consequently, we have quantified
the concentration of all HEPES moieties. Furthermore, the net charge on
each of the four HEPES components are known: $\mathrm{I}: +2$;
$\mathrm{II}: +1$; $\mathrm{III}: 0$; $\mathrm{IV}: -1$. Thereby we have
obtained a specification for the $\mathrm{pH}$ dependent net charge on
HEPES given its total concentration and given the three named
$\mathrm{p}K$ values.  Therefore, with SID denoting strong ion
difference[@Ring2022] and $K_w$ the auto-dissociation of water, the new
charge-balance model for HEPES is:

@@

>>>>

<!--
NOTE: bookdown features (e.g. cross-references) used below
This requires the use of a bookdown output format, e.g.
bookdown::pdf_document2 or bookdown::html_document2
-->

@@

It has been shown[@Ring2022] that $\mathrm{pH}$ in arbitrary fluids can be
understood based on the three fundamental principles of electroneutrality,
conservation of mass, and rules of dissociation as devised from physical
chemistry. It has also been shown[@Ring2022] that the charge-balance approach
allows definition of the pH-dependent buffer-capacity, and provides a definition
of acid content as titratable acidity, based on previous work by
Kildeberg[@Kildeberg1983]. Recently it has been reiterated[@Ring2023-charge]
that it is absolutely necessary for this modeling approach to take into account
the strong ions, which are those assumed to be completely dissociated at any pH.
In contrast to the physico-chemical model of Stewart[@Stewart1978], the
charge-balance model (CBM, henceforth) has no separation between independent and
dependent variables. Under the CBM, all that can be known about the system is
gathered in the expression of charge in an explicitly stated chemical system.
Consequently, if, as an example, $\mathrm{pH}$ is known, and charge is a known
deterministic function of the chemical specification, inverse inferences on
unknown variables can, at least sometimes, be obtained by root finding, since we
know that the charge must be zero. In simple systems, such as a single
monovalent weak acid and strong ions, a single unknown variable is explicitly
estimable[@Ring2022].

Based on explicit description of the chemistry, an expression of the charge of a
given solution can always be specified. It is an interesting question if this
allows _in silico_ modeling of any physiological problem and if every charge
expression has a physiological interpretation. Another question which remains
unanswered is how CBM could allow identification of $[\mathrm{H}^{+}]$ at
several orders below its purported main determinants such as strong ions in the
presence of unavoidable measurement errors in the determinants much larger than
$[\mathrm{H}^{+}]$[@Seifter2014].  Actual sensitivity analysis of data generated
from CBM indicates that say 10% variation in $[\mathrm{H}^{+}]$ has similar
impact on the calculated charge as 10% variation in SID components or
$\mathrm{p}\mathrm{CO_2}$, and the success with reproducing the measured
$\mathrm{pH}$ based on concrete measurements from a large sample of intensive
care unit patients also supports the notion[@Ring2016]. But still, these results
are in need for a real explanation. The more mundane question focused here is if
CBM is rich enough to allow inverse inference in more complex systems such as
with multiprotic buffers.

This question arose under experiments with HEPES buffered sulfate containing
fluids[@Ring2023-acid] in which HEPES was modeled as a monovalent buffer with
pKa $7.5$[@Ring2010] since it was found that estimated $\mathrm{pH}$ was much
too low when observed $\mathrm{pH}$ was less than about $6$. HEPES is a
zwitterion with formula 2-[4-(2-Hydroxyethyl)piperazin-1-yl]ethane-1-sulfonic
acid and structure as shown in \@ref(fig:hepes).

@@

Wikipedia informs[@WP:HEPES] that HEPES has two $\mathrm{p}K$ values, one at $3$
and one at $7.5$, but the reference given[@Johnson2014] in Wikipedia states that
"3" comes from the sulfonate group and $7.5$ from one of the piperazine
nitrogens. At $20^\circ\mathrm{C}$ Good et al[@Good1966] has pKa $3$ and $7.55$
without detailing which groups are at work, while another study[@AlGhobashy2014]
mentions that $3$ and $7.5$ both belong to piperazine amines and informs that
the sulfonate group is dissociated over "almost the whole $\mathrm{pH}$ range".
Furthermore, it has been stated that the nitrogen at the $\mathrm{OH}$ end of
HEPES is protonated in methanol and the nitrogen in the sulfonate end is
protonated in water[@Sledz2010]. This same reference stated that $\mathrm{p}K_a$
of the sulfonic group was between $1$ and $3$ but Wikipedia[@WP:Sulfonic_acid]
has it that sulfonic acids have $\mathrm{p}K_a$ values of $-2.8$ to $-1.9$.

Hence it was not clear how best to model the $\mathrm{pH}$-dependent charge on
HEPES to estimate $\mathrm{pH}$ based on known chemistry. Furthermore, it was
desired to make clear if the $\mathrm{p}K$ values for HEPES could be elucidated
based on measured $\mathrm{pH}$ in fluids with known composition. Normally, the
$\mathrm{p}K$ values are defined by titration but given the apparent
disagreement for HEPES it was of interest to see if the CBM offered another
avenue for understanding.

Herein, these questions will be approached by first developing a model for the
mean charge on HEPES as a function of $\mathrm{pH}$ and applying this to
proposed values of sets of underlying pKs using data obtained
previously[@Ring2023-acid]. Next, it will be examined if data generated with
specified $\mathrm{p}K$ values over extended or restricted range of
$\mathrm{pH}$ values allows recovery of these same $\mathrm{p}K$ values by
non-linear regression. Accepting that negative pH values would hardly be
realized in biological fluid and, anyway, would require dramatic ionic strengths
making conventional understanding of pH modeling problematic, the consequence
for modeling of assuming the first $\mathrm{p}K_1$ fixed at different low values
will be examined. In so far as the developed model for pH dependent charge on
HEPES implies a strict ordering of pK values, as will be seen in the Methods
section, it is not clear if the free pK values can be recovered from in silico
data without such order imposed in the non-linear regression, so this will be
examined.

To demonstrate how binding is the nature of CBM with the proposed model of
charge on HEPES, noise free in silico data was next generated with fixed
$\mathrm{p}K_1$ to examine the effect of pH coverage relative to $\mathrm{p}K_2$
and $\mathrm{p}K_3$ as used in the simulations.

Finally, normal, Cauchy and uniform noise were imposed on $\mathrm{pH}$ to
examine the sensitivity of the estimates to noise ratio and noise form, number
of replications and realized $\mathrm{pH}$. Regarding the last point,
uncertainty in _measured_ $\mathrm{pH}$ is determined by many independent
factors with a probably combined standard-deviation of $0.02$ [@Meinrath2000] at
neutral pH and higher uncertainty at low and high $\mathrm{pH}$ values
[@Leito2002].

@@

<!--
A dash! A dash! My kingdom for a dash!
(The things that rmarkdown users have to do without \usepackage{mhchem})
-->

$\mathrm{I: H^{+}N_1\text-HEP\text-N_4H^{+}SO_3H}$ $\sim \mathrm{p}K_{a1}$ on
sulfonic group

@@

So, we can see that $\mathrm{IV}$ is identified from total HEPES
concentration at a given $\mathrm{pH}$ if the three $\mathrm{p}K$ values
are known, and the derivation clearly requires that $\mathrm{p}K_{a1} <
\mathrm{p}K_{a2} < \mathrm{p}K_{a3}$. Consequently, we have quantified
the concentration of all HEPES moieties. Furthermore, the net charge on
each of the four HEPES components are known: $\mathrm{I}: +2$;
$\mathrm{II}: +1$; $\mathrm{III}: 0$; $\mathrm{IV}: -1$. Thereby we have
obtained a specification for the $\mathrm{pH}$ dependent net charge on
HEPES given its total concentration and given the three named
$\mathrm{p}K$ values.  Therefore, with SID denoting strong ion
difference[@Ring2022] and $K_w$ the auto-dissociation of water, the new
CBM model for HEPES is:

@@

Results
=======

Observed data consisted of $107$ measurements of $\mathrm{pH}$ from $4.63$ to
$8.23$ with HEPES from $48.9 \mathrm{mM}$ to $50 \mathrm{mM}$ and SID from
$–1 \mathrm{mM}$ to $40 \mathrm{mM}$.

```{r predicted-reference, fig.cap = 'Predicted-reference plots for various ways of modelling the pH data', echo = FALSE}
readRDS('observed_final_250124.rds')
```
